---
title: "argList"
author: 
- name: "Sigbert Klinke" 
  email: sigbert@hu-berlin.de
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
vignette: > 
  %\VignetteIndexEntry{plot.matrix} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\usepackage[utf8]{inputenc} 
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```
```{r setup, include = FALSE}
library('argList')
```

# Introduction

When I develop a package, my package functions contain calls to other functions. I want to allow users to pass parameters to these other functions through parameter lists. Normally you would use the ellipsis (`...`) for this. However, this becomes complicated if parameters are to be passed to multiple functions.

```{r, eval=FALSE}
pkgfun <- function(x, y, ...) { 
  fun1(x, ...)
  fun2(y, ...)
}
```

The problem is how to assign the parameters from the ellipsis to the functions `fun1` and `fun2`. One solution is to use parameter lists in `pkgfun`:

```{r, eval=FALSE}
pkgfun <- function(x, y, param1=list(a=1, cex=1), param2=list(b=2, cex=1.5), ...) { 
  param1$x <- x
  do.call(fun1, param1)
  param2$y <- y
  do.call(fun2, param2)
}
```

In addition, it should be possible to use **global** parameters to set all parameters in the parameter lists that support this simultaneously if they have the same name:

```{r, eval=FALSE}
pkgfun(x, y, param1=list(a=1, cex=2), param2=list(b=2, cex=2))
pkgfun(x, y, cex=2)        # alternative call
```

And it should be possible to set individual parameters of a parameter list with **local** parameters:

```{r, eval=FALSE}
# instead of calling
pkgfun(x, y, param2=list(b=2, cex=3))
pkgfun(x, y, param2.cex=3) # alternative call
```

And finally, if there are special parameters, it should be possible to specify their values using the name of the parameter list:

```{r, eval=FALSE}
# instead of calling
pkgfun(x, y, param2=list(b=3, cex=1))
pkgfun(x, y, param2=3)     # alternative call where `b` can be set by `param2`
```

# Installation

The package can be installed from CRAN

```{r, eval=FALSE}
install.packages('argList')
```

or the development version from GitHub

```{r, eval=FALSE}
# install.packages('devtools') # if necessary
devtools::install_github('sigbertklinke/argList')
```

# An example

In a function `test` we call a `plot` command and two `axis` commands:

```{r, fig.width=3, fig.height=3}
test <- function(x, y, ...) {
  plot(x, y, type="n", axes=FALSE)
  axis(side=1, at=pretty(x))
  axis(side=2, at=pretty(y))  
  box()
}

test(iris[,1], iris[,2])
```

**Step 1:** We add three parameter lists with default values

```{r, fig.width=3, fig.height=3}
test <- function(x, y, plot=list(type="n", axes=FALSE, xlab="xlab", ylab="ylab"), 
                 axis1=list(side=1), axis2=list(side=2), ...) {
  plot$x <- x
  plot$y <- y
  do.call('plot', plot)
  axis1$at <- pretty(x)
  do.call('axis', axis1)
  axis2$at <- pretty(y)
  do.call('axis', axis2)
  box()
}

test(iris[,1], iris[,2])
```

**Step 2:** We specify which parameters of the functions `plot` and `axis` can be controlled globally.

```{r}
setParamList('plotparams', graphics::plot.default, deny=c('...', 'x', 'y', 'axes'))
setParamList('axisparams', graphics::axis, deny=c('...', 'at', 'labels'))
# now have look which parameters are allowed
getParamList()
```

If `fun` is provided (not `NULL`), its formal arguments are obtained using `formals(fun)`. It's important to note that the first parameter of `setParamList` serves as an identifier for a distinct parameter set. This allows you to define multiple sets of parameters that can be used in different function call of the same function.

**Step 3:** We modify our function such that **local** and **global** parameters are taken into account.

```{r, fig.width=3, fig.height=3}
test <- function(x, y, plot=list(type="n", axes=FALSE, xlab="xlab", ylab="ylab"),
                 axis1=list(side=1), axis2=list(side=2), ...) {
  args <- update(argList(match.call(), formals(), axis1.side='axis1', axis2.side='axis2'),
                 plot='plotparams', axis1='axisparams', axis2='axisparams')
  #
  args[['plot']]$x <- x
  args[['plot']]$y <- y
  do.call('plot', args[['plot']])
  args[['axis1']]$at <- pretty(x)
  do.call('axis', args[['axis1']])
  args[['axis2']]$at <- pretty(y)
  do.call('axis', args[['axis2']])
  box()
  invisible(args)
}

# define possible global parameter
setParamList('plotparams', graphics::plot.default, deny=c('...', 'x', 'y', 'axes'))
setParamList('axisparams', graphics::axis, deny=c('...', 'at', 'labels'))
# 
test(iris[,1], iris[,2])
# propagate `main` to `plot$main`
test(iris[,1], iris[,2], main="plain")
# and propagate `axis2` to `axis2$side`
test(iris[,1], iris[,2], axis2=4, main="axis2==4")
# `axis2.side` has priority over `axis2=4`
test(iris[,1], iris[,2], axis2=4, axis2.side=2, main="axis2.side=2")
```

The argument list is transformed by:

* `axis1.side='axis1'`: If a parameter `axis1` exists and is not a list then value of `axis1` is propagate to the local variable  `axis1.side`. The parameter `axis1` is set to its default `axis1=list(side=1)`.
* `axis2.side='axis2'`: If a parameter `axis2` exists and is not a list then value of `axis2` is propagate to the local variable  `axis2.side`. The parameter `axis2` is set to its default `axis2=list(side=2)`.
* `plot='plotparams'`: For the `plot` argument 
* is checked if a global parameter is available and added to the parameter list for `plot` and
* any local variable is added to the parameter list for `plot`.
* the same is done for `axis1` and `axis2`.

You may a look to the differences of the arguments in each step: 

```{r}
test <- function(x, y, plot=list(type="n", axes=FALSE, xlab="xlab", ylab="ylab"),
                 axis1=list(side=1), axis2=list(side=2), ...) {
  arg0 <- argList(match.call(), formals())
  arg1 <- argList(match.call(), formals(), axis1.side='axis1', axis2.side='axis2')
  arg2 <- update(arg1, plot='plotparams', axis1='axisparams', axis2='axisparams')
  # compare the modification at the argument lists
  cat(diff(arg0, arg1, 
           xtitle="argList(match.call(), formals())", 
           ytitle="argList(match.call(), formals(), axis1.side='axis1', axis2.side='axis2')"))
  cat(diff(arg1, arg2, 
           xtitle="argList(match.call(), formals(), axis1.side='axis1', axis2.side='axis2')", 
           ytitle="update(., plot='plotparams', axis1='axisparams', axis2='axisparams')"))
}

# define possible global parameter
setParamList('plotparams', graphics::plot.default, deny=c('...', 'x', 'y', 'axes'))
setParamList('axisparams', graphics::axis, deny=c('...', 'at', 'labels'))
str(getParamList())
test(iris[,1], iris[,2], axis2=4, axis2.side=2, main="axis2.side=2")
```

# The algorithm

We differentiate between three categories of variables:

* __Global variables__, which can be present in multiple function calls
* __Argument lists__ for a function's invocation and
* __Local variables__, which overwrite elements from the argument lists'"

Let us assume that in a function `main`, we call another function `fun`. The function `main` has an argument list `funparam=list(...)` allowing us to call `fun` within `main` using `do.call(fun, funparam)`.

* __Step 1:__ We use `match.call()` to obtain the argument list as it was provided to `main`. The `formals()` function retrieves the default arguments of `main`. We then construct an argument list for `main` by adding any missing default argument from `main` to the argument list obtained from `match.call()`.
* __Step 2:__ If parameters other than `match.call()` and `formals()` in `argList` are not a list, they are defined as additional local variables. The parameter will be replaced by its default argument as defined in `formals()`.
* __Step 3:__ Global variables will be added to the argument lists.
* __Step 4:__ All local variables are added to their argument lists.

The priority of the variables is:

1. If a local variable is available their value will be taken. This does depend how the local variable was created, by propagation, user call, or function definition.
2. If no local variable is available the values from the user call are used.
3. Otherwise the default argument list is used.

# Troubleshooting and FAQs

__Is the representation unambiguous?__  
No, you might run into trouble, if you use a dot for other parameter names, e.g. in

`main <- function(axis1.side=4, axis1=list(side=2))`

# Appendix
Additional datasets and code snippets.






